---
title: "post-clustering curation with lulu"
author: "Frédéric Mahé"
date: '2023-03-19'

output:
  rmarkdown::html_document:
    theme: lumen
    toc: yes
    toc_float: TRUE
    keep_md: yes
    # code_folding: hide
---


```{r setup, include=FALSE}
## remove all objects and remaining lulu log files before starting
rm(list = ls())
file.remove(list.files(path = ".",
                       pattern = "lulu.log_*",
                       full.names = TRUE))
knitr::opts_chunk$set(echo = TRUE)
```

***

#### install lulu if need be

devtools is also a requirement

```{r}
## check if lulu is available, if not install it
```


#### load required packages

```{r packages, message=FALSE}
library(lulu)
library(dplyr)
```


## basic usage

The best way to learn a new tool is to play with it!


### simplest example

```{r}
otutable <- data.frame(
    row.names = c("A", "B"),
    s1 = c(9, 1),
    s2 = c(9, 1))

otutable
```

'A' and 'B' co-occur and 'B' is less abundant than 'A'.

```{r}
matchlist <- data.frame(x = "B" , y = "A", z = 99.0)

matchlist
```

'B' has a 99% similarity with 'A'.


Let's see what lulu does with that:

```{r}
lulu::lulu(otutable, matchlist)
```

First, we got a warning indicating that `lulu` uses `funs()`, a
deprecated `dplyr` feature. That feature has been marked as deprecated
in YYYY and will be removed from dplyr in YYYY.

We will focus on the `curated table`:

```{r}
lulu::lulu(otutable, matchlist)$curated_table
```


## issues

- does not work with only one sample,
- does not work with empty matchlists (should output the same table),
- does not work with empty OTUs (no reads, it should skip the OTU),
- no versionning,
- uses deprecated dplyr features,
- no maintenance,
- slow


### partial overlap

```{r}
matchlist_name <- data.frame(x = "B" , y = "A", z = 99.0)
otutable_name <- data.frame(
    row.names = c("A", "B"),
    s01 = c(0, 1), # <= 'B' present, 'A' absent
    s02 = c(9, 1),
    s03 = c(9, 1),
    s04 = c(9, 1),
    s05 = c(9, 1),
    s06 = c(9, 1),
    s07 = c(9, 1),
    s08 = c(9, 1),
    s09 = c(9, 1),
    s10 = c(9, 1),
    s11 = c(9, 1),
    s12 = c(9, 1),
    s13 = c(9, 1),
    s14 = c(9, 1),
    s15 = c(9, 1),
    s16 = c(9, 1),
    s17 = c(9, 1),
    s18 = c(9, 1),
    s19 = c(9, 1),
    s20 = c(9, 1),
    s21 = c(9, 1),
    s22 = c(9, 0)) # <= 'A' present, 'B' absent

otutable_name
```

Note that parent 'A' is absent in one sample where 'B is present', and
present in one sample where 'B' is absent (same total spread). The
relative co-occurence is 20 / 21 = 0.95238, which is greater than
0.95, the default threshold value.


```{r}
## bug: no merging with default parameters
## (minimum_ratio = 0)
lulu::lulu(otutable_name, matchlist_name)
```


***

```{r}
sessionInfo()
```

```{r cleanup, include=FALSE}
file.remove(list.files(path = ".",
                       pattern = "lulu.log_*",
                       full.names = TRUE))
rm(list = ls())
```
